Towards Verifying the Bitcoin-S Library

### 2019-07-03 paper/paper.tex
changes to paper, factor out project setup

%\usepackage{newtxtext}
%\usepackage{newtxmath}
\usepackage[zerostyle=b,scaled=.88]{newtxtt}

@@

\begin{abstract}
  We try to verify properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a subset of Scala
  called \emph{Pure Scala}. We first try to verify the property that
  regular transactions do not create new money. It turns out that
  there is too much code involved that lies outside of the supported
  fragment to make this feasible. However, in the process we uncover
  and fix a bug in bitcoin-s. We then turn to a much simpler (and less
  interesting) property: that adding zero satoshis to a given amount
  of satoshis yields the given amount of satoshis. Here as well a
  significant part of the relevant code lies outside of the supported
  fragment. However, after a series of equivalent transformations we
  arrive at code that we successfully verify.

@@

\paragraph{The Bitcoin-S Library.} The bitcoin-s library is an
implementation of parts of the Bitcoin protocol in Scala
\cite{BitcoinS:website,BitcoinS:github}. In particular, it allows to
serialize, deserialize, sign and validate transactions. The library
uses immutable data structures and algebraic data types but is not
written with formal verification in mind. According to the website,
the library is used in production, handling significant amounts of
cryptocurrency each day \cite{BitcoinS:website}.

@@

\begin{figure}
\begin{lstlisting}[style=scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}
	\caption{Factorial program with specification}
	\label{fig:factorial}
\end{figure}
Our program does not satisfy the specification. An overflow in the
32-bit Int leads to a negative result for the input 17, as Stainless
reports in Figure~\ref{fig:failed}. Changing the type from Int to
BigInt will result in a successful verification.

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
can not generate new coins. We call it the \emph{no-inflation
  property}. Trying to verify it, we uncover and fix a bug in the
bitcoin-s library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. So we turn to a simpler property to verify. The
simplest possible property we can think of is the fact that adding
zero satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{addition-with-zero property} and we try
to verify it in Section~3. Here as well we see that a significant part
of the code lies outside of the supported fragment. However, after a
series of equivalent transformations we arrive at code that we
successfully verify.

@@

\paragraph{A Bug in the checkTransaction function.} Note lines 16 and
17 of the checkTransaction function in Figure~\ref{fig:checktrans}.

@@

\paragraph{Extracting the relevant Code} The relevant code for the
addition of satoshis is in two files: \texttt{CurrencyUnits.scala} and
\texttt{NumberType.scala}.

@@

@@

\section{Project Setup}

\paragraph{Using the Stainless-sbt-plugin vs Using the Stainless-JAR}

We can either use the sbt plugin or a JAR file to check code with Stainless.

More about sbt on https://www.scala-sbt.org/1.x/docs/index.html (features are also listed there)
Installation see Stainless Doc 'Installing Stainless' section

Invoking the JAR on our source code Stainless will verify it.  If we
have a bigger project, this becomes really tricky, because we must
pass all files needed including the dependencies.  This is in contrast
to the sbt plugin, where we can integrate Stainless in our compilation
process.  When we call compile, Stainless verifies the code and stops
the compilation if the verification fails.

Having a static version configured in the sbt build file, every
developer has the same Stainless features available.  This should
prevent incompatibility with new or deprecated features when we use
different plugins.

So the sbt plugin has clear advantages over the JAR file since its
integrated directly.  We do not have to download it manually and find
the right version and if we bump the version we can just edit it in
the build file and every developer is on the same version again.

However, currently there are some drawbacks.  For example the sbt
plugin does not always report errors.
use sbt-plugin because of library management and replicability (everyone uses the same Stainless version configured in sbt)

We use the jar for everything.

Use JAR:
Download all library dependencies including dependencies of dependencies, pass libraries + source to JAR

Use sbt-plugin: integrate in sbt build tool with addSbtPlugin() method

Why JAR:

no sbt because

-- version conflicts, stainless 0.1 uses older sbt version than bitcoin-s (v0.2 not)
-- no output maybe \href{https://github.com/epfl-lara/stainless/issues/484}{issue 484 on
  GitHub}

\paragraph{Reproducing our Results.}

Prerequisites:
\begin{enumerate}
\item Java 8 in /usr/lib/jvm/java-8-openjdk (Stainless requires and Scala recommends Java SE Development Kit 8. Newer Java versions won't work.)
\end{enumerate}

\begin{lstlisting}[language=bash]
 clone repo
 run ./code/bin/stainless.sh <path to code>
   eg. ./code/bin/stainless.sh code/addition/src/main/scala/addition/reduced/
\end{lstlisting}

\section{Extracted Code}
\label{sec:extracted}

\subsection{NumberType.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/reduced/number/NumberType.scala}

\subsection{CurrencyUnits.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/reduced/currency/CurrencyUnits.scala}

@@

\section{Verified Code}
\label{sec:verified}

\subsection{NumberType.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/verified/number/NumberType.scala}

\subsection{CurrencyUnits.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/verified/currency/CurrencyUnits.scala}

>>>>

% latin modern, no more bitmap fonts in pdf
\usepackage{lmodern}
%\usepackage{newtxtext}
%\usepackage{newtxmath}
\usepackage[zerostyle=b,scaled=.88]{newtxtt}

@@

\begin{abstract}
  We try to verify properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a subset of Scala
  called \emph{Pure Scala}. We first try to verify the property that
  regular transactions do not create new money. It turns out that
  there is too much code involved that lies outside of the supported
  fragment to make this feasible. However, in the process we find and
  fix a bug in bitcoin-s. We then turn to a much simpler property:
  that adding zero satoshis to a given amount of satoshis yields the
  given amount of satoshis. Here as well a significant part of the
  relevant code lies outside of the supported fragment. We perform a
  series of equivalent transformations on it until we arrive at code that we
  successfully verify. In that process we find and fix a second bug in
  bitcoin-s.

@@

\paragraph{The Bitcoin-S Library.} The bitcoin-s library is an
implementation of parts of the Bitcoin protocol in Scala
\cite{BitcoinS:website,BitcoinS:github}. In particular, it allows to
serialize, deserialize, sign and validate Bitcoin transactions. The library
uses immutable data structures and algebraic data types but is not
written with formal verification in mind. According to the website,
the library is used in production, handling significant amounts of
cryptocurrency each day \cite{BitcoinS:website}.

@@

\begin{figure}
\begin{lstlisting}[style=scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}
	\caption{Factorial program with specification}
	\label{fig:factorial}
\end{figure}
Our function does not satisfy the specification. An overflow in the
32-bit Int leads to a negative result for the input 17, as Stainless
reports in Figure~\ref{fig:failed}. Changing the type from Int to
BigInt will result in a successful verification.

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
can not generate new coins. We call it the \emph{no-inflation
  property}. Trying to verify it, we uncover and fix a bug in the
bitcoin-s library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. So we turn to a simpler property to verify. The
simplest possible property we can think of is the fact that adding
zero satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{addition-with-zero property} and we try
to verify it in Section~3. Here as well we see that a significant part
of the code lies outside of the supported fragment. We perform a
  series of equivalent transformations on it until we arrive at code that we
  successfully verify. In that process we find and fix a second bug in
  bitcoin-s.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 16 and
17 of the checkTransaction function in Figure~\ref{fig:checktrans}.

@@

\paragraph{Extracting the Relevant Code} The relevant code for the
addition of satoshis is in two files: \texttt{CurrencyUnits.scala} and
\texttt{NumberType.scala}.

@@

\paragraph{A Bug in the checkTransaction Function.}

@@

% \section{Extracted Code}
% \label{sec:extracted}

% \subsection{NumberType.scala}
% \lstinputlisting[style=scala]{../code/addition/src/main/scala/reduced/number/NumberType.scala}

% \subsection{CurrencyUnits.scala}
% \lstinputlisting[style=scala]{../code/addition/src/main/scala/reduced/currency/CurrencyUnits.scala}

@@

% \section{Verified Code}
% \label{sec:verified}

% \subsection{NumberType.scala}
% \lstinputlisting[style=scala]{../code/addition/src/main/scala/verified/number/NumberType.scala}

% \subsection{CurrencyUnits.scala}
% \lstinputlisting[style=scala]{../code/addition/src/main/scala/verified/currency/CurrencyUnits.scala}
